---
title:  'XGBoost Algorithm'
subtitle: 'INFO 523'
author:
- name: Student -  [Gayathri Renganathan, Mohit Rakesh Taparia]
  affiliation: [iSchool]
- name: Instructor -  Cristian Román-Palacios
  affiliation: School of Information, University of Arizona, Tucson, AZ
tags: [R, RStudio, Final Project - XGBoost Algorithm]
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r message=FALSE}
library(readxl)
library(dplyr)
library(caret)
library(xgboost)
library(ggplot2)
library(GGally)
library(fastDummies)
library(corrplot)
library(pROC)
```

**The case study references the Telecom customer churn dataset**
<https://archive.ics.uci.edu/dataset/563/iranian+churn+dataset>

The **read_excel()** method loads the dataset stored in the Excel
format.

```{r warning=FALSE, error=FALSE, message=FALSE}
file_path <- "Data/Customer Churn.csv"
customerInfo <- read.csv(file_path,  sep = ",", header = TRUE)
```

```{r include=TRUE, warning=FALSE}
head(customerInfo,10)
```

```{r warning=FALSE, error=FALSE, message=FALSE}
summary(customerInfo)
```

### **Checking for null Values**

```{r warning=FALSE, error=FALSE, message=FALSE}
sum(is.na(customerInfo))
```

### **Count of Number of customer who got churned or not**

```{r warning=FALSE, error=FALSE, message=FALSE}
customerInfo %>% group_by(Churn) %>% summarise(n=n()) %>% arrange(desc(n))
```

### Plotting churn count in the dataset.

```{r warning=FALSE, error=FALSE, message=FALSE}
ggplot(customerInfo, aes(Churn, fill = factor(Churn))) +
  geom_bar(width = 0.6) + 
  scale_fill_manual(values = c("#FFA07A", "#20B2AA"), 
        labels = c("0" = "Not Churned", "1" = "Churned")) +    
  labs(
    fill = "Churn Status",
    title = "Churn Count Plot",
    x = "Churn",
    y = "Customer Count"
  ) +
  scale_x_continuous(breaks = seq(0, max(customerInfo$Churn), by = 1)) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = "top",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )

```

#### ***Observation1:***

Using this Count plot, target variable is checked for data imbalance (if
there are more number of churn customers than customers who wont churn
or vice versa) There is some data imbalance, as the number of churn
consumers is significantly lower than the number of customers that will
not churn\
Count plot of Age Group to see the distribution of customers across Age

```{r warning=FALSE, error=FALSE, message=FALSE}
ggplot(customerInfo, aes(x = Age.Group, fill = factor(Age.Group))) +
  geom_density(aes(y = ..count..), alpha = 0.7, color = "black", size = 0.7) +
  scale_fill_brewer(palette = "Set2") +  
  labs(
    x = "Age Group",
    y = "Customer Count",
    title = "Density Plot of Age Group Distribution",
    fill = "Age Group"
  ) +
  theme_minimal() +  # Apply a clean, minimal theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"), 
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = "right",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

#### ***Observation2:***

There are more customer in the middle age group (2 and 3)\
Count plot of Age Group to see the distribution of customers across Age

```{r warning=FALSE, error=FALSE, message=FALSE}
ggplot(customerInfo, aes(x = Charge..Amount, fill = factor(Charge..Amount))) +
  geom_bar() + 
  scale_fill_brewer(palette = "Set3") +
  geom_text(stat='count', aes(label=..count..), vjust=-0.5, size=3.5) +
  labs(x = "Charge Amount", y = "Count", fill = "Charge Amount", title = "Distribution of Customers by Charge Amount") +
  scale_x_continuous(breaks = seq(0, max(customerInfo$Charge..Amount), by = 1)) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )
```

#### ***Observation3:***

Charge amount is an ordinal attribute with lowest value being 0 and
highest being 9 The charge amount of 0 has the highest count, and the
count decreases with the charge amount.\
Distribution of Churn count across age group

```{r warning=FALSE, error=FALSE, message=FALSE}
ggplot(customerInfo, aes(x = Age.Group, fill = factor(Churn))) + 
  geom_bar(position = "dodge", color = "black") + 
  scale_fill_manual(values = c("#00AFBB", "#E7B800"), 
                    labels = c("0" = "Not Churned", "1" = "Churned")) +   
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.3, position = position_dodge(0.9), size = 3.5) +
  labs(x = "Age Group", y = "Count", fill = "Churn", title = "Churn Distribution Across Age Groups") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), 
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    legend.position = "right",  
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  ) 
```

#### ***Observation 4:***

Customer in the Middle age group(2 and 3) are churned more compared to
other age group\
Distribution of Churn count across status

```{r warning=FALSE, error=FALSE, message=FALSE}
status_churn_pie <- status_churn_df %>%
  group_by(Status) %>%
  mutate(Percentage = Count / sum(Count) * 100,
         ChurnLabel = ifelse(Churn == 1, "Churned", "Not Churned"),
         StatusLabel = ifelse(Status == 1, "Active", "Inactive"))

ggplot(status_churn_pie, aes(x = "", y = Percentage, fill = ChurnLabel)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  facet_wrap(~StatusLabel, ncol = 2) +
  labs(x = NULL, y = NULL, fill = "Churn", title = "Pie Chart of Churn Distribution Across Status") +
  theme_minimal() +  
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), 
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text = element_blank(),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    panel.grid = element_blank()
  ) +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), size = 3.5) +
  scale_fill_discrete(labels = c("Churned", "Not Churned"))
 
```

#### ***Observation 5:***

Pie Chart Active and Pie Chart Inactive: \
Pie Chart Active: Represents the proportion of different churn statuses
for the first category within the Status variable. For example, if
Status includes "Active" and "Inactive," Pie Chart 1 could represent the
churn distribution for "Active" customers. \
Pie Chart Inactive: Represents the proportion of different churn
statuses for the second category within the Status variable. Following
the same example, Pie Chart 2 would show the churn distribution for
"Inactive" customers.\
Inactive customers are more likely to be churned, compared to the Active
customers.\

Pair plot of all features

```{r warning=FALSE, error=FALSE, message=FALSE}
ggpairs(
  customerInfo[,c("Call..Failure", "Complains", "Subscription..Length", "Charge..Amount","Seconds.of.Use","Frequency.of.use","Age.Group", "Status")],
  upper = list(continuous = wrap("cor", size = 3)), 
  lower = list(continuous = wrap("smooth", size = 0.3)),  
  diag = list(continuous = wrap("densityDiag", alpha = 0.5)), 
  axisLabels = "show" 
) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 2),  
    axis.text.y = element_text(size = 1),  
    panel.grid.minor = element_blank(),  
    panel.grid.major = element_line(color = "gray90"), 
    strip.text = element_text(size = 3, face = "bold")
  )

```

#### ***Observation 6:***

Frequency_of_use and seconds_of_use is strongly correlated.\

Using correlation matrix to find the correlation

```{r warning=FALSE, error=FALSE, message=FALSE}
cor_matrix <- cor(customerInfo)
corrplot(
  cor_matrix,
  method = "pie",           
  type = "upper",           
  tl.col = "black",         
  tl.srt = 45,              
  tl.cex = 0.9,             
  cl.cex = 0.8,            
  number.cex = 0.7,         
  pie.cex = 1.2,          
  mar = c(0, 0, 1, 0),     
  diag = FALSE            
)
```

#### ***Observation 7:***

There is a strong positive correlation with Frequency_of_use and
Seconds_of_use There is a positive correlation between complains and
customer Churn

Dropping the correlated columns to avoid collinearity Dropping Age
column, as `Age` and `Age.Group` represent same feature.

```{r warning=FALSE, error=FALSE, message=FALSE}
customerInfo <- customerInfo %>% select(-c(Frequency.of.use,Age))
```

### **Scaling Continuous Variable**

```{r warning=FALSE, error=FALSE, message=FALSE}
customerInfo_int <- customerInfo[,-c(2,8,9,10,12)]
customerInfo_int <- data.frame(scale(customerInfo_int))
```

### **Creating factors for Categorical Data**

```{r warning=FALSE, error=FALSE, message=FALSE}
customerInfo$Complains <- factor(customerInfo$Complains)
customerInfo$Age.Group <- factor(customerInfo$Age.Group)
customerInfo$Tariff.Plan <- factor(customerInfo$Tariff.Plan)
customerInfo$Status <- factor(customerInfo$Status)
customerInfo$Age.Group <- factor(customerInfo$Age.Group)
customerInfo_scaled <- customerInfo %>% 
                  select(c(2,8,9,10,12)) %>%
                  cbind(customerInfo_int)
```

### **Creating Dummy variable for Categorical Data**

```{r warning=FALSE, error=FALSE, message=FALSE}
sparse_matrix <- model.matrix(Churn ~ ., data = customerInfo_scaled)[,-1]
customerInfo_final <-  customerInfo_scaled %>% select(c("Churn")) %>% cbind(sparse_matrix)

```

### **Split Train and Test Data**

```{r warning=FALSE, error=FALSE, message=FALSE}
target <- "Churn"
#customerInfo_final$Churn <- factor(customerInfo_final$Churn, levels = c(0,1))
features <- setdiff(names(customerInfo_final), target)
set.seed(123)
trainIndex <- createDataPartition(customerInfo_final[[target]], p = 0.8, 
                                  list = FALSE, 
                                  times = 1)
dataTrain <- customerInfo_final[trainIndex, ]
dataTest <- customerInfo_final[-trainIndex, ]
xTrain <- dataTrain[, features]
yTrain <- dataTrain[[target]]
xTest <- dataTest[, features]
yTest <- dataTest[[target]]

```

### **Put our testing & training data into two separates Dmatrixs objects**

```{r warning=FALSE, error=FALSE, message=FALSE}
dtrain <- xgb.DMatrix(data = as.matrix(xTrain), label= yTrain)
dtest <- xgb.DMatrix(data = as.matrix(xTest), label= yTest)
```

## **Logistic regression**

```{r warning=FALSE, error=FALSE, message=FALSE}
logistic_model <- glm(Churn ~ ., data = dataTrain, family = "binomial")
logistic_predict_test <- predict(logistic_model , xTest, type = "response")
predictions_Logistic <- data.frame(actual = yTest,
                          predicted = ifelse(logistic_predict_test > 0.5, 1, 0))
confusionMatrix(as.factor(predictions_Logistic$predicted), as.factor(yTest),positive = "1",  mode = "everything")
```

### **Running with default params**

```{r warning=FALSE, error=FALSE, message=FALSE}

default_param<-list(
        objective = "binary:logistic",
        eval_metric = "logloss",
        eta=0.3, #default = 0.3
        gamma=0,
        max_depth=6, #default=6
        min_child_weight=1, #default=1
        subsample=1,
        colsample_bytree=1)
        
xgbcv <- xgb.cv( params = default_param, data = dtrain, nrounds = 500, nfold = 5, showsd = T, stratified = T, print_every_n = 40, early_stopping_rounds = 10, maximize = F)
```

```{r warning=FALSE, error=FALSE, message=FALSE}

xgb_mod <- xgb.train(data = dtrain, params=default_param, nrounds = 38)
preds_def <- predict(xgb_mod, dtest)
predictions_def_XGB <- data.frame(actual = yTest,
                          predicted = ifelse(preds_def > 0.5, 1, 0))
confusionMatrix(as.factor(predictions_def_XGB$predicted), as.factor(yTest),positive = "1",  mode = "everything")
```

### **Setting up a grid that tunes both these parameters, and also the ETA  (learning rate).**

```{r warning=FALSE, error=FALSE, message=FALSE}
xgb_grid = expand.grid(
nrounds = 1000,
eta = c(0.1, 0.05, 0.03),
max_depth = c(2, 3, 4),
gamma = 0,
colsample_bytree=1,
min_child_weight=c(1, 2, 3),
subsample=1
)
```

### **The next step is to let caret find the best hyper-parameter values  (using 5 fold cross validation).**

```{r warning=FALSE, error=FALSE, message=FALSE}
set.seed(27042018)
my_control <-trainControl(method="cv", number=5)
#xgb_caret <- train(x=xTrain, y=yTrain, method='xgbTree', trControl= my_control, tuneGrid=xgb_grid) 
#xgb_caret$bestTune
```

### **Forming the Params based on the best tune**

```{r warning=FALSE, error=FALSE, message=FALSE}
best_tune_param<-list(
        objective = "binary:logistic",
        eval_metric = "logloss",
        eta=0.05, #default = 0.3
        gamma=0,
        max_depth=4, #default=6
        min_child_weight=3, #default=1
        subsample=1,
        colsample_bytree=1
)
```

### **Cross validation to determine the best number of rounds  (for the given set of parameters).**

```{r warning=FALSE, error=FALSE, message=FALSE}
xgbcv <- xgb.cv( params = best_tune_param, data = dtrain, nrounds = 700, nfold = 5, showsd = T, stratified = T, print_every_n = 40, early_stopping_rounds = 10, maximize = F)
```

```{r warning=FALSE, error=FALSE, message=FALSE}
xgb_mod <- xgb.train(data = dtrain, params=best_tune_param, nrounds = 352)
```

```{r warning=FALSE, error=FALSE, message=FALSE}
preds <- predict(xgb_mod, dtest)
predictions_XGB <- data.frame(actual = yTest,
                          predicted = ifelse(preds > 0.3, 1, 0))
head(predictions_XGB)
```

## **Confusion matrix for the predictions**

```{r warning=FALSE, error=FALSE, message=FALSE}
confusionMatrix(as.factor(predictions_XGB$predicted), as.factor(yTest),positive = "1", mode = "everything")
```

### **Plotting the roc curve**

```{r warning=FALSE, error=FALSE, message=FALSE}
plot.roc(yTest, as.numeric(predictions_XGB$predicted), 
         print.auc = TRUE, 
         col = 'red', 
         lwd = 2, 
         main = "Enhanced ROC Curves", 
         legacy.axes = TRUE, 
         print.auc.y = 0.7, 
         print.auc.x = 0.6, 
         print.auc.cex = 1.2)
plot.roc(yTest, as.numeric(predictions_def_XGB$predicted), 
         print.auc = TRUE, 
         col = 'blue', 
         lwd = 2, 
         add = TRUE,  
         print.auc.y = 0.6, 
         print.auc.x = 0.6, 
         print.auc.cex = 1.2)
plot.roc(yTest, as.numeric(predictions_Logistic$predicted), 
         print.auc = TRUE, 
         col = 'black', 
         lwd = 2, 
         add = TRUE,  
         print.auc.y = 0.5, 
         print.auc.x = 0.6, 
         print.auc.cex = 1.2)
legend("bottom", 
       legend = c("XGBoost - Fine Tuned", "XGBoost", "Logistic"),
       lty = c(1, 1), 
       lwd = c(2, 2, 2), 
       col = c("red", "blue", "black"), 
       cex = 0.9, 
       box.lty = 0, 
       inset = 0.05)
```

## ***Feature Importance***

```{r}
#view variable importance plot
mat <- xgb.importance (feature_names = colnames(xTrain),model = xgb_mod)
xgb.ggplot.importance(importance_matrix = mat, rel_to_first = TRUE, measure = "Gain")
```

### Distribution of Churn count across Complains

```{r}
ggplot(customerInfo, aes(x = Complains, fill = factor(Churn))) + 
  geom_bar(position = "dodge", color = "black") +  # Grouped bar plot with black border
  scale_fill_manual(values = c("#00AFBB", "#E7B800")) +  # Custom colors for the Churn groups
  scale_x_discrete(labels = c("0" = "No complains", "1" = "Complains")) +
  geom_text(stat = 'count', aes(label = ..count..), vjust = -0.3, position = position_dodge(0.9), size = 3.5) +  # Adding count labels
  labs(x = "Complains", y = "Count", fill = "Churn", title = "Churn Distribution Across Complains") +
  theme_minimal() +  # Apply a minimalistic theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  # Center and bold the title
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    legend.position = "right",  # Position the legend on the right
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  ) 
```

### Distribution of Churn count across Seconds.of.use

```{r}
ggplot(customerInfo, aes(x = Churn, y = Seconds.of.Use, fill = factor(Churn))) + 
  geom_boxplot() +  
  scale_x_discrete(labels = c("0" = "Not Churned", "1" = "Churned")) +
  labs(x = "Churn", y = "Seconds.of.use", fill = "Churn", title = "Churn Distribution Across Seconds.of.use") +
  theme_minimal() +  # Apply a minimalistic theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  # Center and bold the title
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    legend.position = "right",  # Position the legend on the right
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  ) 
```

## ***Final Observation*** 

From Model Inactive Customers (Status) are most likely to be churned.
Customers who use for less seconds are likely to be churned. Customers
who raised Complaints are churned compared to customers with no
complaints.
