---
title:  'XGBoost Algorithm'
subtitle: 'INFO 523'
author:
- name: Student -  [Gayathri Renganathan, Mohit Taparia]
  affiliation: [iSchool]
- name: Instructor -  Cristian Román-Palacios
  affiliation: School of Information, University of Arizona, Tucson, AZ
tags: [R, RStudio, Final Project - XGBoost Algorithm]
output: html_document

---

```{r}
library(readxl)
library(dplyr)
library(caret)
library(xgboost)
library(ggplot2)
library(GGally)
```

The case study references the Telecom customer churn dataset from the EMPRESS-i on https://archive.ics.uci.edu/dataset/563/iranian+churn+dataset

The read_excel() method loads the dataset stored in the Excel format. 

```{r}
file_path <- "Data/Customer Churn.csv"
customerInfo <- read.csv(file_path,  sep = ",", header = TRUE)
```


```{r}
head(customerInfo,10)
```

```{r}
summary(customerInfo)
```

Checking for null Values
```{r}
sum(is.na(customerInfo))
```

Count of Number of customer who got churned or not
```{r}
customerInfo %>% group_by(Churn) %>% summarise(n=n()) %>% arrange(desc(n))
```

Plotting churn count in the dataset.
```{r}
ggplot(customerInfo, aes(Churn, fill = factor(Churn))) +
  geom_bar()+ 
  labs(fill = "Churn", title = "Churn count plot")
  
```
Observation1: 
using this Count plot,  target variable is checked for data imbalance  (if there are more number of churn customers than customers who wont churn or vice versa)
There is some data imbalance, as the number of churn consumers is significantly lower than the number of customers that will not churn


Count plot of Age Group to see the distribution of customers across Age
```{r}
ggplot(customerInfo, aes(Age.Group, fill = factor(Age.Group))) +
  geom_bar()+ 
  labs(fill = "Age Group", title = "Age Group count plot")
  
```
Observation2: 
There are more customer in the middle age group (2 and 3)


Count plot of Age Group to see the distribution of customers across Age
```{r}
ggplot(customerInfo, aes(Charge..Amount, fill = factor(Charge..Amount))) +
  geom_bar()+ 
  labs(fill = "Charge Amount", title = "Charge Amount count plot")
  
```
Observation3: 
Charge amount is an ordinal attribute with lowest value being 0 and highest being 9
The charge amount of 0 has the highest count, and the count decreases with the charge amount.


Distribution of Churn count across age group
```{r}
customerInfo %>% 
  ggplot(aes(x = Age.Group, fill = factor(Churn))) + 
  geom_bar()  
```
Observation:
Customer in the Middle age group(2 and 3) are churned more compared to other age group


Distribution of Churn count across status 
```{r}
customerInfo %>% 
  ggplot(aes(x = Status, fill = factor(Churn))) + 
  geom_bar(position = "dodge") +
  scale_x_discrete(limits = c(1,2))
```
Observation:
Status is ordinal attribute; 1 - Active ; 2 - Inactive
Inactive customers are more likely to be churned, compared to the Active customers.

Pair plot of all features
```{r}
ggpairs(customerInfo[,c("Call..Failure", "Complains", "Subscription..Length", "Charge..Amount","Seconds.of.Use","Frequency.of.use","Age.Group", "Status")])
```
Observation:

Frequency_of_use and seconds_of_use is strongly correlated.

Using correlation matrix to find the correlation
```{r}
library(corrplot)

cor <- cor(customerInfo)
corrplot(cor, method="pie")
```
Observation:
 
 There is a positive correlation with Frequency_of_sms and customer_value
 
 

